Lensing in the Schwarzschild Spacetime Applicable to Neutron Stars

### 2022-09-01 src/tex/ms.tex
NS Boosting Calculated.

For a null geodesic, we also have the normalization condition 
\begin{equation}
\label{normalization}
g_{\mu \nu} p^\mu p^\nu = 0.
\end{equation}
Exploiting the rotational symmetry of the $\theta$-$\phi$ subspace, we can,
without loss of generality, work in the equatorial plane, considering only
geodesics which have $\chi^\theta (\lambda) = \pi /2$.  In this case, Equation
\ref{normalization} becomes
\begin{equation}
-\frac{e^2}{1 - \frac{2}{\chi^r}} + \frac{l^2}{(\chi^r)^2} + \frac{1}{1-\frac{2}{\chi^r}} \left( \frac{\dd \chi^r}{\dd \lambda} \right)^2 = 0.
\end{equation}
Rescaling the affine parameter, we obtain
\begin{equation}
\label{radial-path}
\frac{\dd \chi^r(s)}{\dd s} = \pm \sqrt{1 - \frac{b^2\left(1 - \frac{2}{\chi^r(s)} \right)}{(\chi^r(s))^2}},
\end{equation}
with $b \equiv l/e$ called the ``impact parameter''\footnote{$b$ is so named
because it gives the distance at infinity between the geodesic and the nearest
parallel line running radially from the center of the star.} of the trajectory.
This equation and Equations \ref{eq:conserved-e} and \ref{eq:conserved-l}
completely determine a trajectory given $l$ and $e$.

@@

Expanding to leading order in $1/\rNS$, we obtain 
\begin{equation}
    \chi^\phi\left( \rNS \right) = \arcsin x + \frac{2 \left( 1 - \sqrt{1-x^2} \right)}{x \rNS} + \ldots.
\end{equation}
We see that the grazing ray ($x = 1$) impacts the stellar surface at longitude 
\begin{equation}
    \phiNS = \frac{\pi}{2} + \frac{2}{\rNS} + \ldots = \frac{\pi}{2} + \frac{\rSchw}{\rNS} + \ldots
\end{equation}

@@

To be continued\ldots.

>>>>

For a null geodesic, we also have the normalization condition 
\begin{equation}
\label{normalization}
g_{\mu \nu} p^\mu p^\nu = 0.
\end{equation}
Exploiting the rotational symmetry of the $\theta$-$\phi$ subspace, we can,
without loss of generality, work in the equatorial plane, considering only
geodesics which have $\chi^\theta (\lambda) = \pi /2$.  In this case, Equation
\ref{normalization} becomes
\begin{equation}
    \label{eq:pr-normalization}
-\frac{e^2}{1 - \frac{2}{\chi^r}} + \frac{l^2}{(\chi^r)^2} + \frac{1}{1-\frac{2}{\chi^r}} \left( \frac{\dd \chi^r}{\dd \lambda} \right)^2 = 0.
\end{equation}
Rescaling the affine parameter, we obtain
\begin{equation}
\label{radial-path}
\frac{\dd \chi^r(s)}{\dd s} = \pm \sqrt{1 - \frac{b^2\left(1 - \frac{2}{\chi^r(s)} \right)}{(\chi^r(s))^2}},
\end{equation}
with $b \equiv l/e$ called the ``impact parameter''\footnote{$b$ is so named
because it gives the distance at infinity between the geodesic and the nearest
parallel line running radially from the center of the star.} of the trajectory.
This equation and Equations \ref{eq:conserved-e} and \ref{eq:conserved-l}
completely determine a trajectory given $l$ and $e$.

@@

Expanding to leading order in $1/\rNS$, we obtain 
\begin{equation}
    \label{eq:chi-phi-impact}
    \chi^\phi\left( \rNS \right) = \arcsin x + \frac{2 \left( 1 - \sqrt{1-x^2} \right)}{x \rNS} + \ldots.
\end{equation}
We see that the grazing ray ($x = 1$) impacts the stellar surface at longitude 
\begin{equation}
    \phiNS = \frac{\pi}{2} + \frac{2}{\rNS} + \ldots = \frac{\pi}{2} + \frac{\rSchw}{\rNS} + \ldots
\end{equation}

It will be useful to have the momentum of the trajectory at the point of impact,
when $\chi^r = \rNS$, $\chi^\theta = \pi/2$, and $\chi^\phi$ given in Eq.\
\eqref{eq:chi-phi-impact}.  From Eq.\ \eqref{eq:conserved-e}, we see that 
\begin{equation}
    p^t = \frac{e}{1 - \frac{2}{\rNS}} \simeq e \left( 1 + \frac{2}{\rNS} + \ldots \right);
\end{equation}
from Eq.\ \eqref{eq:conserved-l}, and replacing the specific angular momentum
$l$ with the impact parameter $b = l/e$, we see that 
\begin{equation}
    p^\phi = \frac{b e}{\rNS^2} = \frac{x e}{\rNS} \left( 1 + \frac{1}{\rNS} + \ldots \right);
\end{equation}
and from Eq.\ \eqref{eq:pr-normalization}, we see that 
\begin{equation}
    p^r = e \sqrt{1 + \left( \frac{2 + 3 \rNS}{\rNS^3} - 1 \right) x^2} \simeq e \sqrt{1-x^2} \left( 1 + \frac{3 x^2}{2 \left( 1 - x^2 \right) \rNS^2} + \ldots \right).
\end{equation}

For a ray that does not approach the surface in the equatorial plane, but rather
strikes the surface at a colatitude $\theta_0$ (but still with impact parameter
$b = x \bNS$), the tangent vector $\phi$ component rotates into $\theta$ and
$\phi$ components, and the tangent is 
\begin{eqnarray}
    \label{eq:pt-final}
    p^t & \simeq & e \left( 1 + \frac{2}{\rNS} + \ldots \right) \\ 
    \label{eq:pr-final}
    p^r & \simeq & e \sqrt{1-x^2} \left( 1 + \ldots \right) \\
    \label{eq:ptheta-final}
    p^\theta & \simeq & \frac{x e \cos \theta_0}{\rNS} \left( 1 + \frac{1}{\rNS} + \ldots \right) \\
    \label{eq:pphi-final}
    p^\phi & \simeq & \frac{x e}{\rNS} \left( 1 + \frac{1}{\rNS} + \ldots \right),
\end{eqnarray}
where we have kept only terms $\order{1/\rNS}$ beyond leading order.

@@

Plugging in Eqs.\ \eqref{eq:pt-final}, \eqref{eq:pr-final},
\eqref{eq:ptheta-final}, and \eqref{eq:pphi-final}, we obtain the relation
between the energy observed at infinite distance ($e$) and the emitted energy at
the surface ($e_\mathrm{surf}$) for a ray with impact factor $b = x \bNS$
hitting the surface of the neutron star at colatitude $\theta_0$:
\begin{multline}
    \frac{e}{e_\mathrm{surf}} = 1 + y \vRot \sin \theta_0 + \left( y^2 - \frac{1}{2} \right) \vRot^2 \sin^2 \theta_0 \\ 
    - \frac{1}{\rNS} \left( 1 - \left( y^2 - \frac{1}{2} \right) \vRot^2 \sin^2 \theta_0 \right) + \ldots, 
\end{multline}
where $y = x \sin \theta_0$ is the ``equatorial'' component of the
(dimensionless) impact factor.
